GPU Efficient Texture Based Bezier Curve Evaluation

### 2015-07-18 Paper/paper.tex
Progress: Spell check and some other things.

\begin{abstract}
\small
Modern graphics techniques expose internal parameters to allow re-use and to help seperate the technical implementation from artistic usage cases.  A popular choice is to expose parameters as customizable curves and to quantize the curves into textures.  Quantization leads to either lower quality results, or more texture memory being used to accomodate higher sampling frequencies.  The technique presented in this paper leverages the capabilities of GPU texture samplers to allow more efficient storage and evaluation of both integral and rational Bezier curves of any order, resulting in higher fidelity for the same costs.  Piecewise curves, B-Splines and Nurbs are addressed, and there are also limited applications towards vector graphics.
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=5in]{Figure2.png}
  \caption{\textit{Left:} A quadratic Bezier curve written to the green channel, and the baked out equivelant using 4 pixels written to the red channel.  \textit{Right:} The technique from this paper, also using 4 pixels, written to the green channel, and the baked out equivelant written to the red channel again for comparison.}
  \label{fig:quickcomparison}
\end{figure}

@@

A quick comparison of the visual quality of these three techniques can be seen in \autoref{fig:quickcomparison}.

@@

The core of this technique is that the $N$-dimensional linear texture interpolation capabilities on the GPU can be mathematically equivelant to De Casteljeau's algorithm for evaluating Bezier curves.  We can use linear texture sampling on textures of increasing dimensionalities to have the texture sampler calculate points on curves of increasing order.

@@

One dimensional linear texture sampling is just the linear interpolation between two data points.  This is equivelant to the De Casteljeau algorithm when evaluating a linear curve of order 1.  In both cases, it is just linear interpolation between two values.

@@

To further see how bilinear interpolation can be equivelant to evaluating a quadratic Bezier curve with the De Casteljeau algorithm, \autoref{fig:decasteljeauBilinear} shows it visually, and you can also compare the GLSL source code in \autoref{lst:GLSLCurves}.

@@

  \begin{figure}
    \begin{tikzpicture}[x=12.5cm,y=4.25cm]
      % border
      \draw (0cm,0cm) -- (12.5cm,0cm) -- (12.5cm,4.25cm) -- (0cm,4.25cm) -- (0cm, 0cm);   
      % raw texture, bilinear sampled texture, curve results
        \node[anchor=south west,inner sep=0] at (0.125cm,0.125cm) {\includegraphics[width=4cm,height=4cm]{Figure4Texture.png}};
        \node[anchor=south west,inner sep=0] at (4.25cm,0.125cm) {\includegraphics[width=4cm,height=4cm]{Figure4Bilinear.png}};
        \node[anchor=south west,inner sep=0] at (8.375cm,0.125cm) {\includegraphics[width=4cm,height=4cm]{Figure4Curves.png}};
        % representative line for where the quadratic bezier curve lives
        \draw[yellow,->,line width=0.5mm]    (5.25cm,1.125cm) -- (7.25cm,3.125cm);    
        \draw (0.08,0.25) node[] {\Huge{A}};
        \draw (0.25,0.25) node[] {\Huge{B}};
        \draw (0.08,0.75) node[] {\Huge{B}};
        \draw (0.25,0.75) node[] {\Huge{C}};            
    \end{tikzpicture}
    \caption{\textit{Left:} a 2x2 texture storing control points for a quadratic Bezier curve in each color channel.  \textit{Middle:} The same texture as viewed when using bilinear texture sampling.  The yellow line indicates where texture samples are taken from to evaluate the quadratic Bezier curve. \textit{Right:} The curves resulting from sampling along the yellow line (Alpha curve ommited).}   
    \label{fig:texbilcurve}
  \end{figure}

@@

The De Casteljeau algorithm is equivelant to the Bernstein form of Bezier curves.

@@

That they are equivelant means that the Bernstein form for a given $n$ must also be able to be evaluated by $n$-Linear interpolation.  A linear curve should be able to be evaluated by linear interpolation, a quadratic curve should be able to be evaluated by bilinear interpolation, a cubic curve should be able to be evaluated by trilinear interpolation, and so on.

@@

    \begin{tikzpicture}
      \draw[step=1cm,gray,very thin] (0,0) grid (7,1);
      \draw (0.5,0.5) node {$A$};
      \draw (1.5,0.5) node {$B$};
      \draw (2.5,0.5) node {$C$};
      \draw (3.5,0.5) node {$D$};
      \draw (4.5,0.5) node {$E$};
      \draw (5.5,0.5) node {$F$};
      \draw (6.5,0.5) node {$G$};     
    \end{tikzpicture}      
    \caption{\textit{Top:} A cubic curve with control points $A,B,C,D$ encoded in a 1d texture with a size of 4 pixels.  \textit{Middle:} Two piecewise curves encoded in a 1d texture.  The control points for the first curve are $A,B,C,D$ and the control points for the second curve are $D,E,F,G$.  \textit{Bottom:} With $C0$ continuitiy, a redundant $D$ can be removed to store these curves in 7 pixels instead of 8. }    
    \label{fig:texlayeout1d}
  \end{figure}

@@

This allows us to either calculate a quadratic bezier curve with a single texture read, or allows us to get the quadratic level values needed to make a higher order curve.

@@

Removing that redunant row, it means that we start with a $2$x$2$ texture for a quadratic curve, and then just add a row of pixels for each order above quadratic.  That makes the size of our texture required be only $(2,N)$.

@@

\begin{lstlisting}[caption={GLSL for evaluating a cubic curve encoded in a $(2,4)$ pixel 2d texture.  Bilinear texture sampling used to evaluate the first two levesl of the De Casteljeau algorithm, then the process is continued both with the De Casteljeau algorithm, as well as the Bernstein form of a linear Bezier curve (lerp).}, label={lst:GLSLCubicTexture2D}]
// 2x4 2d texture, rounded to the nearest powers of 2 from 2x3.
uniform sampler2D uSampler; 
const vec2 c_textureSize = vec2(2.0,4.0);

@@

Using three dimensional textures (often refered to as volumetric textures) allows the texture interpolator to evaluate trilinear interpolation, which allows calculations of 3rd order (cubic) Bezier curves (\autoref{fig:decdimensionality}).

This allows us to either calculate a cubic bezier curve with a single texture read, or allows us to get the cubic level values needed to make a higher order curve.

@@

You can combine two cubic curves to make a quartic curve, or three cubic curves to make a quintic curve.  If you were to put $2$x$2$x$2$ cubic curves back to back, you would notice a redundancy like we saw in lower dimensionality textures.  Instead of going that route though, we can just build on the redundancy removement from two dimensional textures.

@@

You could also set it up such that R,G,B,A were meant to be combined together to make a quintic (degree 5) curve.  You'd only get a single curve value per texture read, but it would be 3 degrees higher.  You could combine the color channels either with De Casteljeaus algorithm, or with the cubic Bernstein Bezier form.

If you you are combining M color channels, each of which are values from curves of order N, the resulting curve value will be order N+M-1.  The inutitive explanation to this, is that each channel has the ability to add one more control point to the curve, regardless of what texture dimensionality you are working with.

@@

B-splines can also be converted to piecewise Bezier curves using Boehm's algorithm, so even though this technique only explicitly deals with Bezier curves, B-splines could be convered to piecewise Bezier curves and could be used in that form.

@@

Rational Bezier curves are useful because they can be used to make shapes which cannot be formed from integral Bezier curves, including perfectly representing conic sections.  Rational Bezier curves are defined by the equation:

@@

Looking closely, we can see that we essentially have one Bezier curve devided by another.  The numerator has a weighting and a control point, while the denominator has only a weighting, but if we were to multiply the weighting into the control point in the numerator, we would essentially have one Bezier curve divided by another.

@@

So far we have only talked about one dimensional - or explicit - bezier curves which take the form of $y=f(x)$.  Multidimensional Bezier curves are evaluated seperately per axis though, which means that you can encode information per axis in different color channels.

@@

Another problem with using this technique for vector graphics, is that since the texture sampler is doing at least some of the calculations for us, we don't have access to the values of the control points, which can make some operations harder.  Luckily, one property of Bezier curves is that affine transformations done to control points is equivelant to affide transformations done to points on the curve.  This means that we can do affine transformations on the curve points provided to us by the texture sampler, and it will be as if we had performed them on the control points themselves.

@@

The table below was created by running a 1000x1000 pixel shader program where each pixel evaluated a curve using either linear texture sampling (HW or hardware evaluation), or did multiple texture reads to get the control points and then calculated the curve values inside of the shader program (SW or software shader program evaluation).  Actual performance characteristics in real world scenarios will be based on other factors, such as whether the texture is in the texture cache, or whether you are texture fetch bound or compute bound in the shader program.

@@

There may also be usefulness in curves in polar form to aproximate lobes for lighting algorithms.

>>>>

\begin{abstract}
\small
Modern graphics techniques expose internal parameters to allow re-use and to help separate the technical implementation from artistic usage cases.  A popular choice is to expose parameters as customizable curves and to quantize the curves into textures.  Quantization leads to either lower quality results, or more texture memory being used to accommodate higher sampling frequencies.  The technique presented in this paper leverages the capabilities of GPU texture samplers to allow more efficient storage and evaluation of both integral and rational Bezier curves of any order, resulting in higher fidelity for the same costs.  Piecewise curves, B-Splines and NURBS are addressed, and there are also limited applications towards vector graphics.
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=5in]{Figure2.png}
  \caption{\textit{Left:} A quadratic Bezier curve written to the green channel, and the baked out equivalent using 4 pixels written to the red channel.  \textit{Right:} The technique from this paper, also using 4 pixels, written to the green channel, and the baked out equivalent written to the red channel again for comparison.}
  \label{fig:quickcomparison}
\end{figure}

@@

A quick comparison of the visual quality of these three techniques can be seen in \autoref{fig:quickcomparison}.  Besides being useful for exposing internal parameters, the technique presented in this paper is also useful as an alternative to traditional shader look up tables which use linear interpolation to get points between data points.

@@

The core of this technique is that the $N$-dimensional linear texture interpolation capabilities on the GPU can be mathematically equivalent to De Casteljeau's algorithm for evaluating Bezier curves.  We can use linear texture sampling on textures of increasing dimensionalities to have the texture sampler calculate points on curves of increasing order.

@@

One dimensional linear texture sampling is just the linear interpolation between two data points.  This is equivalent to the De Casteljeau algorithm when evaluating a linear curve of order 1.  In both cases, it is just linear interpolation between two values.

@@

To further see how bilinear interpolation can be equivalent to evaluating a quadratic Bezier curve with the De Casteljeau algorithm, \autoref{fig:decasteljeauBilinear} shows it visually, and you can also compare the GLSL source code in \autoref{lst:GLSLCurves}.

@@

  \begin{figure}
    \begin{tikzpicture}[x=12.5cm,y=4.25cm]
      % border
      \draw (0cm,0cm) -- (12.5cm,0cm) -- (12.5cm,4.25cm) -- (0cm,4.25cm) -- (0cm, 0cm);   
      % raw texture, bilinear sampled texture, curve results
        \node[anchor=south west,inner sep=0] at (0.125cm,0.125cm) {\includegraphics[width=4cm,height=4cm]{Figure4Texture.png}};
        \node[anchor=south west,inner sep=0] at (4.25cm,0.125cm) {\includegraphics[width=4cm,height=4cm]{Figure4Bilinear.png}};
        \node[anchor=south west,inner sep=0] at (8.375cm,0.125cm) {\includegraphics[width=4cm,height=4cm]{Figure4Curves.png}};
        % representative line for where the quadratic bezier curve lives
        \draw[yellow,->,line width=0.5mm]    (5.25cm,1.125cm) -- (7.25cm,3.125cm);    
        \draw (0.08,0.25) node[] {\Huge{A}};
        \draw (0.25,0.25) node[] {\Huge{B}};
        \draw (0.08,0.75) node[] {\Huge{B}};
        \draw (0.25,0.75) node[] {\Huge{C}};            
    \end{tikzpicture}
    \caption{\textit{Left:} a 2x2 texture storing control points for a quadratic Bezier curve in each color channel.  \textit{Middle:} The same texture as viewed when using bilinear texture sampling.  The yellow line indicates where texture samples are taken from to evaluate the quadratic Bezier curve. \textit{Right:} The curves resulting from sampling along the yellow line (Alpha curve omitted).}   
    \label{fig:texbilcurve}
  \end{figure}

@@

The De Casteljeau algorithm is equivalent to the Bernstein form of Bezier curves.

@@

That they are equivalent means that the Bernstein form for a given $n$ must also be able to be evaluated by $n$-Linear interpolation.  A linear curve should be able to be evaluated by linear interpolation, a quadratic curve should be able to be evaluated by bilinear interpolation, a cubic curve should be able to be evaluated by trilinear interpolation, and so on.

@@

    \begin{tikzpicture}
      \draw[step=1cm,gray,very thin] (0,0) grid (7,1);
      \draw (0.5,0.5) node {$A$};
      \draw (1.5,0.5) node {$B$};
      \draw (2.5,0.5) node {$C$};
      \draw (3.5,0.5) node {$D$};
      \draw (4.5,0.5) node {$E$};
      \draw (5.5,0.5) node {$F$};
      \draw (6.5,0.5) node {$G$};     
    \end{tikzpicture}      
    \caption{\textit{Top:} A cubic curve with control points $A,B,C,D$ encoded in a 1d texture with a size of 4 pixels.  \textit{Middle:} Two piecewise curves encoded in a 1d texture.  The control points for the first curve are $A,B,C,D$ and the control points for the second curve are $D,E,F,G$.  \textit{Bottom:} With $C0$ continuity, a redundant $D$ can be removed to store these curves in 7 pixels instead of 8. }    
    \label{fig:texlayeout1d}
  \end{figure}

@@

This allows us to either calculate a quadratic Bezier curve with a single texture read, or allows us to get the quadratic level values needed to make a higher order curve.

@@

Removing that redundant row, it means that we start with a $2$x$2$ texture for a quadratic curve, and then just add a row of pixels for each order above quadratic.  That makes the size of our texture required be only $(2,N)$.

@@

\begin{lstlisting}[caption={GLSL for evaluating a cubic curve encoded in a $(2,4)$ pixel 2d texture.  Bilinear texture sampling used to evaluate the first two levels of the De Casteljeau algorithm, then the process is continued both with the De Casteljeau algorithm, as well as the Bernstein form of a linear Bezier curve (lerp).}, label={lst:GLSLCubicTexture2D}]
// 2x4 2d texture, rounded to the nearest powers of 2 from 2x3.
uniform sampler2D uSampler; 
const vec2 c_textureSize = vec2(2.0,4.0);

@@

Using three dimensional textures (often referred to as volumetric textures) allows the texture interpolator to evaluate trilinear interpolation, which allows calculations of 3rd order (cubic) Bezier curves (\autoref{fig:decdimensionality}).

This allows us to either calculate a cubic Bezier curve with a single texture read, or allows us to get the cubic level values needed to make a higher order curve.

@@

You can combine two cubic curves to make a quartic curve, or three cubic curves to make a quintic curve.  If you were to put $2$x$2$x$2$ cubic curves back to back, you would notice a redundancy like we saw in lower dimensionality textures.  Instead of going that route though, we can just build on what we did for two dimensional textures.

@@

You could also set it up such that R,G,B,A were meant to be combined together to make a quintic (degree 5) curve.  You'd only get a single curve value per texture read, but it would be 3 degrees higher.  You could combine the color channels either with De Casteljeau's algorithm, or with the cubic Bernstein Bezier form.

If you you are combining M color channels, each of which are values from curves of order N, the resulting curve value will be order N+M-1.  The intuitive explanation to this, is that each channel has the ability to add one more control point to the curve, regardless of what texture dimensionality you are working with.

@@

B-splines can also be converted to piecewise Bezier curves using Boehm's algorithm, so even though this technique only explicitly deals with Bezier curves, B-splines could be converted to piecewise Bezier curves and could be used in that form.

@@

\begin{figure}
    \begin{tikzpicture}[x=12cm,y=6cm]
      \node[anchor=south west,inner sep=0] at (0cm,0cm) {\includegraphics[width=12cm,height=6cm]{Rational.png}};  
    \end{tikzpicture}

    \caption{A rational bezier spline showing cosine in green and sine in red.  The white line is the true value of those functions. \textit{Left:} Evaluated with hardware sampling and then division.  \textit{Right:} Evaluated by reading the control points into the shader program, calculating the curves, and then doing a division.}
    \label{fig:rationalbeziersincos}
  \end{figure}

Rational Bezier curves are useful because they can be used to make shapes which cannot be formed from integral Bezier curves, including perfectly representing conic sections which mean that they can also represent sine and cosine as seen in \autoref{fig:rationalbeziersincos}.  Rational Bezier curves are defined by the equation:

@@

Looking closely, we can see that we essentially have one Bezier curve divided by another.  The numerator has a weighting and a control point, while the denominator has only a weighting, but if we were to multiply the weighting into the control point in the numerator, we would essentially have one Bezier curve divided by another.

@@

So far we have only talked about one dimensional - or explicit - Bezier curves which take the form of $y=f(x)$.  Multidimensional Bezier curves are evaluated separately per axis though, which means that you can encode information per axis in different color channels.

@@

Another problem with using this technique for vector graphics, is that since the texture sampler is doing at least some of the calculations for us, we don't have access to the values of the control points, which can make some operations harder.  Luckily, one property of Bezier curves is that affine transformations done to control points is equivalent to affine transformations done to points on the curve.  This means that we can do affine transformations on the curve points provided to us by the texture sampler, and it will be as if we had performed them on the control points themselves.

@@

The table below was created by running a 1000x1000 pixel shader program where each pixel evaluated a curve using either linear texture sampling (HW or hardware evaluation), or did multiple texture reads to get the control points and then calculated the curve values inside of the shader program (SW or software shader program evaluation).  The numbers shown are the time spent rendering each frame, in microseconds and the texture format is R8G8B8A8.  Actual performance characteristics in real world scenarios will be based on other factors, such as whether the texture is in the texture cache, or whether you are texture fetch bound or compute bound in the shader program.

@@

There may also be usefulness in curves in polar form to approximate lobes for lighting algorithms.
